---
title: "Reproducible Research: Peer Assessment 1 - David Lloyd"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r echo=TRUE}
library(stats)
library(dplyr)
activity_data <- read.csv("activity.csv")
activity_data <- filter(activity_data, steps != "NA")
```

## What is mean total number of steps taken per day?


##### Total steps per day

```{r echo=TRUE, fig.height=8, fig.width=10}

total_steps_day <- activity_data %>% 
    group_by(date) %>%
    summarise(total_steps = sum(steps) )

hist(as.numeric(total_steps_day$total_steps),  col="green", xlab = "Date", main = "Total Steps per Day")

```

##### Average and Median steps per day summarized in the table below

```{r echo=TRUE}
print(total_steps_day %>%
    summarise(avg_steps = mean(total_steps), median_step = median(total_steps) )
)
```



## What is the average daily activity pattern?

Compute the average number of steps per time interval

```{r echo=TRUE}
average_steps_interval <- activity_data %>%
    group_by(interval) %>%
    summarise(avg_interval_steps = mean(steps) )
```



Time series plot of the interval and the average number of steps taken across all days

```{r echo=TRUE, fig.height=8, fig.width=10}
with(average_steps_interval, plot(x=interval, y=avg_interval_steps, type = "l", xlab = "5 min interval", ylab = "Average Steps Taken"))
```


The 5-minute interval with the highest average number of steps accross all days

```{r echo=TRUE}
max_steps <- average_steps_interval %>%
  summarise(max_step_interval = max(avg_interval_steps)) 


print( filter(average_steps_interval, avg_interval_steps==max_steps[[1,1]])  %>%
         select(interval)
      )
```
      

## Imputing missing values


This time get all data, not just non-NA

```{r echo=TRUE}
activity_data_full <- read.csv("activity.csv", stringsAsFactors = F)
```


Cursor through each row in the data frame; if the steps are NA in that row, find the corresponding interval step average
computed in average_steps_interval data frame and replace the NA with that interval average

```{r echo=TRUE}
 for(i in 1:nrow(activity_data_full)) {

      if(is.na(activity_data_full[i,1])) {
          activity_row <- filter(average_steps_interval, interval == activity_data_full[i,3])
          activity_data_full[[i,1]] =  activity_row[[1,2]]
      }
 }
```


Total steps per day

```{r echo=TRUE}
total_steps_day <- activity_data_full %>% 
  group_by(date) %>%
  summarise(total_steps = sum(steps) )

```

Average and Median steps per day

```{r echo=TRUE}
print(total_steps_day %>%
        summarise(avg_steps = mean(total_steps), median_step = median(total_steps) )
)
```

#Histogram of Total steps per day
```{r echo=TRUE, fig.height=8, fig.width=10}
hist(as.numeric(total_steps_day$total_steps),  col="green", xlab = "Date", main = "Total Steps per Day")
```


As one might expect, if one replaces NA values with the interval step average the variance/standard deviation decreases
(i.e. the "bell curve" becomes more narrow)


## Are there differences in activity patterns between weekdays and weekends?

First step is adding a 2 level factor variable value indicatating whether or now the date is a weekday or weekend day

```{r echo=TRUE}
activity_data_week <- activity_data %>%
  mutate(weekend = as.factor( case_when( weekdays(as.Date(date, "%Y-%m-%d")) == "Saturday" | 
                              weekdays(as.Date(date, "%Y-%m-%d")) == "Sunday" ~ "weekend",
                              TRUE ~ "weekday" ) ) )
```


Compute averages by weekend/weekday

```{r echo=TRUE}
activity_data_weekend <- filter(activity_data_week, weekend=="weekend")
average_steps_interval_weekend <- activity_data_weekend %>%
  group_by(interval) %>%
  summarise(avg_interval_steps = mean(steps) )

activity_data_weekday <- filter(activity_data_week, weekend=="weekday")
average_steps_interval_weekday <- activity_data_weekday %>%
  group_by(interval) %>%
  summarise(avg_interval_steps = mean(steps) )
```


Panel plot two time series of interval average times on weekends and weekdays

```{r echo=TRUE, fig.height=12, fig.width=10}
par(mfrow=c(2,1))
with(average_steps_interval_weekend, plot(x=interval, y=avg_interval_steps, type = "l",  xlab = "5 min interval", ylab = "Average Steps Taken", main = "Weekend"))

with(average_steps_interval_weekday, plot(x=interval, y=avg_interval_steps, type = "l", xlab = "5 min interval", ylab = "Average Steps Taken", main = "Weekday"))
```

